Metric Spaces with Expensive Distances

### 2019-04-17 paper/expensive-distances-main.tex
Some changes in paper

\section{Experiments on spanners}
We run experiments on the points sampled from the low-dimensional Euclidean space to investigate
experimentally the performance of these heuristics. Clearly, for this metric space, our
cost model is not meaningful since distance comparisons are cheap;
but we picked this environment for controlled experiments.
In order to test the \textsc{BlindQuasiSorted} algorithms we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
By \textit{sparseness} of a graph with $n$
vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.

\begin{figure}[ht]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
        legend pos=outer north east]

@@

\begin{figure}[ht]
        \begin{tikzpicture}
            \begin{axis}[xlabel=\# Points,ylabel=\# Edges / \# Points,
                legend pos=outer north east]

@@

\begin{figure}[ht]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# Diagrams,
        legend pos=outer north east]

@@

\begin{figure}[ht]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# All distances,
        legend pos=outer north east]

@@

We tested the algorithms for $\eps \in \Set{0.01, 0.1, 0.2, 0.5}$ on the following sets of points in dimensions $d = 2,3,4,5$:
\begin{enumerate}
    \item In the \textbf{uniform} test set points are sampled uniformly at
        random from the unit cube in $\mathbb{R}^d$.
    \item In the \textbf{normal} test set points are sampled from the standard
        normal distribution in $\mathbb{R}^d$.
    \item In the \textbf{clustered} test set we first sample cluster centers uniformly 
        at random from $[0,10000]^d$, and then we add normally distributed noise around
        each of the centers. The number of clusters is chosen so that each cluster
        contains 50 points.
    \item The test set \textbf{exp} consists of points of the form $(2^{\xi_1}, \dots, 2^{xi_d})$,
        where $\xi_i$'s are i.i.d. random variables with uniform distribution on $[1,25]$.
\end{enumerate}
In all experiments the algorithms that we tested compared in the same way,
so we only present results for the \textbf{uniform} point set in dimension 2.

From these experiments we conclude that the algorithm with the highest chance
of saving a significant amount of distance computations on real data, like large
persistence diagrams, is \bgrdy. Even though \brndm also substantially reduces
the number of computed distances, its performance is significantly worse;
other algorithms that we tested do not produce sparse spanners.
While it would seem plausible that access to approximate value of the distance 
could be exploited in the spanner construction, 
we could not find a working heuristic; \bqsgrdy and \bqsshaker produce extremely
dense spanners.

Finally, we ran experiments on two realistic data sets,
namely the persistence diagrams of the shapes
from  the McGill shape benchmark \cite{zhang2005mcgill}. 
More precisely, we generated two sets of diagrams,
\dtype{original} and \dtype{perturbed}. The \dtype{original}
diagrams were computed on the points of the 450 original
point clouds from the shape dataset; we only looked
at dimension 0, since the diagrams in this dimension are the largest.
We also added random noise to points of the original shapes,
and computed persistence diagrams in dimension 0 of these modified
shapes; these diagrams form the \dtype{perturbed} dataset.
Persistence diagrams were computed with \libraryname{DIPHA}
library.

@@

The proof is given in the appendix \cref{sec:proof}.

>>>>

\section{Experiments on spanners}

We performed two types of experiments: on points sampled from a low-dimensional Euclidean
space and on persistence diagrams with Wasserstein distance.
Clearly, our model is not meaningful in a Euclidean space, where
distance computations are cheap.
However, in this case we can easily check the performance
of our heuristics on a variety of easily controllable
data sets.  In order to test the \textsc{BlindQuasiSorted} algorithms, we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
By \textit{sparseness} of a graph with $n$
vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.

%We run experiments on the points sampled from the low-dimensional Euclidean space to investigate
%experimentally the performance of these heuristics. Clearly, for this metric space, our
%cost model is not meaningful since distance comparisons are cheap;
%but we picked this environment for controlled experiments.
%In order to test the \textsc{BlindQuasiSorted} algorithms we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
%By \textit{sparseness} of a graph with $n$
%vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.

We tested the algorithms for $\eps \in \Set{0.01, 0.1, 0.2, 0.5}$ on the following sets of points in dimensions $d = 2,3,4,5$:
\begin{enumerate}
    \item In the \textbf{uniform} test set points are sampled uniformly at
        random from the unit cube in $\mathbb{R}^d$.
    \item In the \textbf{normal} test set points are sampled from the standard
        normal distribution in $\mathbb{R}^d$.
    \item In the \textbf{clustered} test set we first sample cluster centers uniformly 
        at random from $[0,10000]^d$, and then we add normally distributed noise around
        each of the centers. The number of clusters is chosen so that each cluster
        contains 50 points.
    \item The test set \textbf{exp} consists of points of the form $(2^{\xi_1}, \dots, 2^{xi_d})$,
        where $\xi_i$'s are i.i.d. random variables with uniform distribution on $[1,25]$.
\end{enumerate}
In addition to the standard Euclidean norm on $\RR^n$, we used also $\ell_1$ and $\ell_\infty$ norms.
In all these experiments the algorithms that we tested compared in the same way,
so we only present results for the \textbf{uniform} point set in dimension 2.

\begin{figure}[!htbp]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
        legend pos=outer north east]

@@

\begin{figure}[!htbp]
        \begin{tikzpicture}
            \begin{axis}[xlabel=\# Points,ylabel=\# Edges / \# Points,
                legend pos=outer north east]

@@

From these experiments we conclude that the algorithm with the highest chance
of saving a significant amount of distance computations on real data is \bgrdy. 
Even though \brndm also substantially reduces
the number of computed distances, its performance is significantly worse;
other algorithms that we tested do not produce sparse spanners.
While it would seem plausible that access to approximate value of the distance
could be exploited in the spanner construction,
we could not find a working heuristic; \bqsgrdy and \bqsshaker produce extremely
dense spanners.

The persistence diagrams that we used as
realistic data sets,
come from the McGill shape benchmark \cite{zhang2005mcgill}.
More precisely, we generated two sets of diagrams,
\dtype{original} and \dtype{perturbed}. The \dtype{original}
diagrams were computed on the points of the 450 original
point clouds from the shape dataset; we only looked
at dimension 0, since the diagrams in this dimension are larger in size.
We also added random noise to points of the original shapes,
and computed persistence diagrams in dimension 0 of these modified
shapes; these diagrams form the \dtype{perturbed} dataset.
Persistence diagrams were computed with \libraryname{DIPHA}
library.

We also compare the time that is needed to compute the matrix of all pairwise distances
naively with the time that is needed to approximate it with \bgrdy spanner.
The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
If one compares the running time with the sparseness results,
it becomes evident that in this case the overhead of \bgrdy algorithm
is rather small, compared to the time needed to compute the distances.
Let us emphasize that the diagrams we have are still rather small, with a typical cardinality
between 100 and 300 points. Nevertheless, even on this data set \bgrdy spanner
provides a substantial improvement in terms of real running time.

\begin{figure}[!htbp]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# Diagrams,
        legend pos=outer north east]

@@

\begin{figure}[!htbp]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# All distances,
        legend pos=outer north east]

@@

@@

The full proof is given in the appendix \cref{sec:proof_ann}.
